Three-dimensional binary-liquid lattice Boltzmann simulation of microchannels with rectangular cross sections

### 2011-09-23 Text/Submission/answers.tex
Almost finished a referee report.

\textbf{Question:}
\begin{quotation}
In Figures 7 and 8, it's better to give the reported curves by Liu and Wang and Hazel and Heil, so
that it's easy to find the agreement is well or not.
\end{quotation}

\textbf{Answer:} We redrew all the figures. For the figure 8 the data was extracted from the
figure and rescaled for comparison reasons.

@@

\textbf{Question:}
\begin{quotation}
The main quantity that the paper focuses on is the thickness of the film between the bubbles and the
wall. As shown by Figure 8 and as mentioned by the Authors in Section 4, the film thickness varies
along the bubble. However, the Authors do not explain the choice of the location of the measurement
of the film thickness. This choice is of primary importance when comparing results with those of
the
literature and in particular with works on finger propagation in which the film thickness is
measured
”at infinity”.
\end{quotation}

\textbf{Answer:} The explanations are added in lines 434-453.

\textbf{Question:}
\begin{quotation}
It may be appropriate to introduce Figure 8 on thickness variation before the main results on ﬁlm
thickness.
\end{quotation}

\textbf{Answer:} We agree and moved the section before the results on the capillary number
dependency.

\textbf{Question:}
\begin{quotation}
The lines 54-55 indicating that ”numerical simulations and experimental studies showed [...]
Reynolds
number effects on the film thickness for capillary numbers larger than 0.003” is in contradiction
with
the lines 32-33 which state ”negligible Reynolds number effects on the film thickness for a
relatively
wide range of Reynolds numbers.”
\end{quotation}

\textbf{Answer:} We elaborate the Reynolds number influence in lines 34-43.

\textbf{Question:}
\begin{quotation}
Although inertia has been shown to have a minor effect on ﬁlm thickness, inertia does influence
some flow variables in the Bretherton’s problem (see Ref. 3 for instance). To check whether inertia
is important or not, and to discuss the assumption of uniform density, the Authors should provide
the reader with the Reynolds numbers associated with each simulation. This is also required for the
consistency of Figure 7 in which results are compared with other works of the literature.
\end{quotation}

\textbf{Answer:} Thank you for pointing this out. We included the discussions of the Reynolds
effects in the answer to the question above. As well, we added Reynolds numbers to the Table for
the capillary range regime and the grid depedency study.

\textbf{Question:}
\begin{quotation}
In this work, the diphasic flow is driven by a body force. The body force is therefore the control
parameter in the simulations. The Authors should give more detail about this body force. Indeed,
although the estimation of the body force from the pressure gradient is well described, it is not
clear
where this body force appears in the governing equations (Equations 2, 3 and 4). Does this body
force correspond to the variable $F_i$ in Equation 2? Should not this force or its equivalent appear
in
the Navier-Stokes equations (Equation 4)?
\end{quotation}

\textbf{Answer:} We updated equation (4) to include the force contribution. As well, the definition
of $P_{\alpha\beta}$ is added. $F_i$ is the force population which corresponds to the force
inclusion $\bm{F}$. We added the citation which describes how the force population is formed and
how it
mimics the force behavior.

\textbf{Question:}
\begin{quotation}
I would suggest to calculate the Bond number to check if the shapes of the bubbles and the main
results of the paper are not inﬂuenced.
\end{quotation}

\textbf{Answer:} 
The maximum Bond number $Bo=\frac{\rho g N_y^2}{\gamma}$ in the
current simulation equals to $7$, which is less than the number for which the shape of the bubble is
affected according to \cite{zheng-large-ratio}.

\textbf{Question:}
\begin{quotation}
The use of the variable $Ca_{lit}$ and the calculation of the corresponding pressure gradient yield
estimations of the body force to apply and of the film thickness to choose grid resolution. The
presentation
of the corresponding procedure is repeated twice in Section 5-1 for $Ca=0.005$ and $Ca=0.05$. It is
again described in a different way in Section 5-4. Similar equations for the calculation of the
pressure
gradient are thus given in Equations 6, 14 and 16. The Authors should summarize these equations
into one unless they explain their differences. In fact, it would probably be straightforward not to
detail the calculation of estimated variables since they are intermediary variables which allow the
design of the benchmark but do not inﬂuence directly the results. Also, the Authors could possibly
say they control and impose directly the body force without describing all the estimation procedure.
\end{quotation}

\textbf{Answer:} Thank you for the suggestion. We agree that there is redundant information
about initialization and this was taken into account by reducing repetitive text and
introducing step-by-step explanations, see Section about initialization (lines $259-305$). We also
added
the
recommendations for controlling parameters in the same section.  The
dimensionalization of the problem and proper benchmarking is quite difficult in the lattice
Boltzmann framework. You can refer to the lbmethod.org/forum where half of all questions related to
the lattice Boltzmann method are refered to the non-dimensionalization procedure. Therefore, we
decided to
keep suggestions for controlling and initialization of simulations as informative guidelines for the
LBM community.

\textbf{Question:}
\begin{quotation}
 Moreover, the estimations of the pressure gradient for the initialisation of the scheme are
obtained
under the assumption of an axisymmetric Poiseuille flow. However, among the ”Key words”, one can
read the key word ”Flow between plates” in agreement with the 2D non axisymmetric simulations
that are done in this work. The Authors should discuss this choice of estimating the pressure
gradient with the axisymmetric assumption instead of a Hele-Shaw flow (factor 1/12 instead of 1/8
in Equation 5).
\end{quotation}

\textbf{Answer:} The factor $1/8$ in Eq. (5) is taken for the maximum velocity for the Poiseille
flow between
plates. Our assumption is that the bubble travels with the constant velocity which equals to the
maximum velocity of the Poiseuille profile between plates. Factor $1/12$ comes from the averaging
of the flow rate. However, bubble moves faster than the liquid, i.e. the average liquid velocity.
That is why the factor $1/8$ was chosen. The simulations show that factor $1/8$ works reasonably
well.

\textbf{Question:}
\begin{quotation}
The Section on LBM is concise and this seems a reasonable choice when publishing such a work in a
journal of applied science. The Authors would improve signiﬁcantly the manuscript by defining more
rigorously some of the variables of the model. For instance, the Authors refer to surface tension in
lines 106, 111 and 123 via the parameter k, the weights w and the quantity 8kA/9 respectively.
A suggestion is to emphasize within an array the links between physical parameters and the model
parameters. The variables $i$, $f_i^{eq}$ , $g_i$ should be deﬁned and the Equations for $f$ and $g$
in Equation
3 should be described (which equation corresponds to continuity equation...). Equation 2 could
perhaps be rewritten with the use of a generic variable in order to remove any ambiguity with the
physical variable f of Equation 3. The notation $i = 1 \div 8$ is unusual, it would be clearer to
use a
classical notation such as $1 \leq i \leq 8$. The sound speed $c_s^2$ should be deﬁned just after
its
introduction in line 107.
\end{quotation}

\textbf{Answer:} We wrote an additional paragraph emphasizing the connection of the
physical parameters with the lattice Botlzmann parameters (lines $149-160$). As well more
explanations are
put towards the
interface surface tension description (lines $136-139$). The reference to the
continuity and the Navier-Stokes equations are added (lines $140-143$). The equation (2) is not
rewritten in terms of
generic variables as far as it's a historical LBM notation for the Navier-Stokes and phase
governing equations. The notations have been corrected.

\textbf{Question:}
\begin{quotation}
The name of the Section ”Lattice Boltzmann benchmark” lets the reader think that this Section
would describe the numerical procedure (assumptions, choice of the geometry and the grid, boundary
conditions, control parameters, initialization of the scheme). This is not the case because this
Section
develops the challenges and the difficulties of using the Lattice Boltzmann method to bubble ﬂows.
This Section should be renamed or reorganized.
\end{quotation}

\textbf{Answer:} We renamed the section to ``Approach to benchmark''.

\textbf{Question:}
\begin{quotation}
The Authors should make a distinction between variables in lattice units and dimensional or non
dimensional variables. For instance, in line 195, the definition $H_{eff} = N_y - 2$ where $N_y$ is
the
number of nodes in the vertical direction indicates that $H_{eff}$ has lattice units. However, this
is in
contradiction with the use of $H_{eff}$ in Equation 5. The Authors should define each variable
without
any ambiguity on its units. Besides, the unit of $U_{bubble}$ should also be given. These
improvements
would fit well in the Subsection named ”The nondimensionalization and initialization procedure”
which does not in fact introduce much nondimensionalisation procedure.
\end{quotation}

\textbf{Answer:} We added a decription paragraph about the approach to units (lines 149-160). In
fact, all equations are the same in the
lattice Boltzmann framework or in a physical framework. One can substitute any quantity in
preferred units as soon as the capillary number is the same in the LBM and the physical world.

\textbf{Question:}
\begin{quotation}
Some notations are unclear. For instance, the film thickness is denoted as $h_{\infty}$ in Equation
7 whereas
it is denoted $\delta$ in most of the paper. The notations w or $\delta_{sim}$ are also found before
line 381 or in
Table 3 respectively. The thickness of the interface is denoted $\xi$ in Figure 1 but the same
variable
is
denoted $5\xi$ in Table 1. The speed of the bubble is denoted $U_{\mathrm{bubble}}$ in some places
and $U_b$ in others.
The variable $Uin$ Equation 5 does not seem to be deﬁned.
\end{quotation}

\textbf{Answer:} We thoroughly corrected all the inconsistencies.

\textbf{Question:}
\begin{quotation}
Three figures in the manuscript are scanned from the litterature (Ref. 3, 10 and 16). This is
unusual
for an article which is not a review. I encourage the Authors to remove at least the two ﬁrst ones
which are not essential.
\end{quotation}

\textbf{Answer:} We eliminated unnecessary pictures and references to them.

\textbf{Question:} 
\begin{quotation}
In Figure 7, the points of data reported from the numerical work of Ref. 3 exhibit a surprisingly
noisy trend that corresponds more to an experimental study. The Authors should explain how they
obtain this data and plot them.
\end{quotation}

\textbf{Answer:} We added a description to the figure caption explaining data acquisition.

\textbf{Question:} 
\begin{quotation}
The plots of Figure 8 are of less quality than the other figures of the paper.
\end{quotation}

\textbf{Answer:} Unfortunately, it was not possible to extract data as a picture from the paper by
\citet{sehgal-microchannel}. Thus, the curves were extracted from the graphs and represented in a
new figure scaled to the present simulations for comparison reasons.

\textbf{Question:}
\begin{quotation}
The work of Yang seems to be close from the present study since it involves the
solution of Bretherton
flow with a Lattice Boltzmann method. The Authors should discuss this paper and emphasize the
differences between their work and this paper.
\end{quotation}

\textbf{Answer:} We outlined the differences in the introduction section (lines $92-100$).

\textbf{Question:} It would be interesting if the Authors could mention a study in which a level-set
method is used in
order to complete the bibliography.

@@

\bibliographystyle{unsrtnat}
\bibliography{answers}

>>>>

\textbf{Question:}
\begin{quotation}
In Figures 7 and 8, it's better to give the reported curves by Liu and Wang and Hazel and Heil, so
that it's easy to find the agreement is well or not.
\end{quotation}
\textbf{Answer:} We added additional figures.

\textbf{Question:}
\begin{quotation}
In Figure 9, the variation of bubble radii in the range of capillary number 0.05<=Ca<=0.6 is
still different from the results of Hazel and Heil and of Liu and Wang. The critical capillary
number of Liu and Wang is close to 0.8 and the critical capillary number of Hazel and Heil is close
to 0.04, but the value of the authors is 0.1. The authors should analyze the reason of the
difference.
\end{quotation}
\textbf{Answer:} The reviewer is right that the critical capillary number of Liu and Wang is close
to 0.8 as it can be seen from their digitized numbers. However, Liu and Wang state in their paper
that the critical capillary number is 0.1. That means they used weaker criterion for the axial
radius to be equal to the diagonal radius. Therefore we consider that the critical number for
transition is between 0.04 and 0.1. The criterion is not well defined and depends on the flow
conditions. For example, Hazel and Heil simulate the air finger propagation with the bubble length
of more than 10 diameters of the channel. In our case, we simulate the bubble train where each
bubble has a certain variation of the film thickness over the bubble length. We chose to measure the
film thickness in the middle of the bubble. There is an assumption that the bubble length is long
enough (3 diameters) for thickness to coincide with the air finger film thicknesses of the infinite
length \cite{cerro-space}. However, in the case of smaller capillary the film thickness variation
is quite large. Thus, more defined criteria should be given in papers in order to validate our
results. Moreover, some formulations like the propagation of the semiinfinite air finger is not
achievable due to computational resources constraints.

\textbf{Question:}
\begin{quotation}
In Figure 10, there is a deviation of the relative velocity between the present and published data.
Can the authors extend the range of capillary numbers as conducted by Liu and Wang?
\end{quotation}
\textbf{Answer:} Unfortunately that requires months with the uniform grids formulations of the
lattice Boltzmann method or some grid refinement needed. We didn't have aim to go towards smaller
capillary numbers as they nicely covered by original work of \citet{bretherton}.

\textbf{Question:}
\begin{quotation}
Check some errors, e.g., on Page 7 Line 137 "0.05>=Ca<=6", on Page 17 Line 300, "82×82×1500 and
82×82", and on Page 24 References 18 and 20.
\end{quotation}
\textbf{Answer:} Thank you for pointing that. All the errors are corrected. Reference 20 is
available online and is not published in journals.

@@

\textbf{Question:}
\begin{quotation}
Generally speaking, the motion of a drop in a channel is a function of channel shape, drop size,
ratio between driving force and surface tension, density ratio, and viscosity ratio.  For the flow
of a long gas bubble (Taylor-Bretherton) the density and viscosity ratios are nearly zero, making it
possible to fit the film thickness as a function of a single dimensionless number Ca.  In the
problem studied in this paper, the bubble is long, the density ratio is unity and the viscosity
ratio is 0.1 (gas over liquid).  As such, while the film thickness in this system is also reduced to
a function of the Capillary number, I do not think that the system considered by the authors is a
Taylor-Bretherton system, as the density and viscosity ratios are different.

The authors used the sensitivity of the film thickness on Re to justify that the sensitivity on
density ratio is low (the first paragraph of Page 6).  However, I do not see the one-to-one
connection between Re and the density ratio.  Are there any direct numerical/experimental evidence
that density ratio has small effect on the film thickness, or evidence that Re is solely (or nearly
solely) determined by the density ratio?
\end{quotation}

\textbf{Answer:}  The reviewer is right about not a full compliance with the classical
Bretherton/Taylor problem. We added more explanations (lines 99-112). The system is governed with
the non-dimensional numbers. Mainly, apart from geometrical dimensions the system can be fully
descisribed by the following set of parameters: Re, Ca and Bo (Eo), where the Bond number Bo is
taken into the account density ratio and is mainly involved in problems for a bubble change.
However, the Bond number in simulations and real experimental setups is low. That's why we consider
it as a non-leading non-dimensional parameter.

\textbf{Question:}
\begin{quotation}
Aside from this critique, it looks like the numerical resolution needs some clarifications.  While
some numbers and discussions were provided, they seem to be in conflict with each other.  For
example, in the discussion given in Page 8 and top of Page 9, the authors stated that the minimum
resolution needed to resolve the film is 600x600x9000, and, if a quarter of the domain is simulated,
a resolution of 300x300x9000 is needed.  However, the domain sizes presented later are all smaller,
e.g. 100x100x1500, 160x160x1500, 160x160x2400, 200x200x3000 (Page 12).  How were these grid
resolutions selected, and what is the effect of grid resolution on the result?
\end{quotation}

\textbf{Answer:} As an example, the corresponding grid $600\times600\times9000$ was taken to resolve
$\delta=0.49 H_{\mathrm{eff}}$ for $Ca=0.03$ from reported simulations of \citet{heil-threedim}.
However, luckily for us for current simulations the value of the film thickness was higher
than $0.49 H_{\mathrm{eff}}$. That's why the computational resources demand is lower and we used
smaller grids. The simulation for $600\times 600 \times 9000$ is too much computationally demanding.
To give account for numbers one simulation with the grid $200\times200\times 3000$ is obtained after
$24$ days on $32$ CPUs. The values of these grids were chosen to resolve the film thickness for the
smallest capillary number (film thickness is twice bigger than the interface thickness). We've done
inextensive grid resolution study (see Fig. 9) to support our results.  We clarified
this issue on lines 168-171.

\textbf{Question:}
\begin{quotation}
A couple of minor issues are listed below:
Page 4, figure 1 - I suppose the vertical axis is $r_d$ and $r_h$ normalized by $H_eff$?  The axis
label perhaps should be clarified.
Page 5, line 18 - missing "from" between "capillary number" and "simulations".
Page 7, line 19 - "0.05 >= Ca <= 6" should be "0.05 <= Ca <= 6".
Page 10 - On the lattice Boltzmann free energy model, what are the boundary conditions for f and g
populations?  Were they simple bounce-back?  Is contact angle implemented in the method?
Page 12, line 13 - "k = A = 0.004, 0.04".  I do not understand this notation - does this mean k is
always equal to A and their values should be between 0.004 and 0.04?
Page 17, line 17-19 - The text says that Figure 9 contains data with different initializations and
different grids.  However, the distinction between data sets cannot be viewed from the figure. 
Perhaps different symbols should be used so that the readers can tell one data set from another?
Page 17, line 40 - "differencies" should be "differences".
Page 19, line 25 - ". the film thickness as twice the interface thickness".  Please see the previous
comment on grid resolution.
Page 20, line 19 - Again, the notation used for the range of A and k is rather unusual.
Page 29, line 46 - ". relaxed after to the axis radius" - perhaps a typo in the sentence?
\end{quotation}

\textbf{Answer:} Thank you for improving the article. All the corrections are taken into the
account. We also added more explanations to the text to address the following questions:
\begin{itemize}
 \item For $f$ and $g$ populations the wall was simulated using the simple bounce-back rule. We
showed before that results do not depend on the wettability of the wall
\cite{kuzmin-binary2d}. Thus, the walls were chosen with neutral wettability. 
 \item We chose $k$ and $A$ equal to each other as this choice conserves the film thickness ($5$
grid spacings). $k=A=0.04$ is taken for most simulations. $k=A=0.004$ is taken for simulation with 
$Ca=6$. Usually for stability issues the choice of $k=A$ is a good one. The values of $k$ and $A$
can vary in different limits. However, the stability is drastically reduced for $k=A=0.25$ or if
$k=A \ll 1$. The corresponding text is added (lines 219-221).
\item Figure 9 shows the results grid independency. It's exactly the reason why we didn't
distinguish different initializations and grid numbers as proper designed simulations should not
depend on it. Corresponding text is added (lines 316-317).
\end{itemize}

\textbf{Question:}
\begin{quotation}
Finally, I have a question on how the initial distributions of f and g were determined?  Did the
simulation start with uniform distributions of f and g populations and they "phase separate" to form
the bubble?
\end{quotation}

\textbf{Answer:} Initial distributions $f$ and $g$ were taken through equilibrium distribution
functions, which were constructed from initial phase and density fields. Density was taken uniform
with the value $1$ across the domain. Bubble was initialized as the rectangular parallelepiped in
the middle domain. After that the system evolves with time and interplay between surface tension
and viscous forces form the bubble. It was shown \cite{kuzmin-binary2d} that the final steady-state
result does not depend on the initialization.

Reviewer III:
\textbf{Question:}
\begin{quotation}
Check some spelling errors, also in the abstract;
\end{quotation}

\textbf{Answer:} We thoroughly corrected all found mistakes.

\textbf{Question:} 
\begin{quotation}
Improve Figure quality: Figs. 4 and above all 5 are hard to read and understand. The ticks should
be the same height, those in Fig. 5 are hardly visible.
\end{quotation}

\textbf{Answer:} We constructed figures with the same size ticks, title and labels. For better
readability we rearranged figures to be of larger size.

\textbf{Question:} 
\begin{quotation}
The bibliography should be improved: when introducing LB, I think that more hints should be given
to the applications, for instance to the case of magnetically-driven vaporization in ferrofluids
studies or regarding multiphase approach based on pseudopotentials.
\end{quotation}

\textbf{Answer:} We expanded the introduction for industry applications and discussion of
multiphase models (lines 117-123).
\textbf{Question:}
\begin{quotation}
Did the Authors perform any comparison with pseudopotential multiphase approaches? I am
especially referring to recent advances of pseudopotentials, such as the case of doubly-attractive
potentials and to recent papers regarding LB multiphase simulation across scales.
\end{quotation}

\textbf{Answer:} Unfortunately, to our best knowledge there is only one study of the pseudopotential
model towards the Bretherton-Taylor phenomenon. The study is in 2D and cannot be applicable to this
case. However we outlined the approaches for pseudopotential models (lines 117-123).

\textbf{Question:}

@@

\bibliographystyle{unsrtnat}
\bibliography{paper}
